Data organization in spreadsheets

### 2017-07-30 manuscript.Rmd
Small changes following review by TAS editor

Murrell [-@murrell2013] contrasts data that is formatted for humans to
view by eye with data that is formatted for a computer. He provides an
extended example of computer code to extract data from a set of files
with complex arrangements. It is important that data analysts be able
to work with such complex data files. But if the initial arrangement
of the data files is planned with the computer in mind, this would
simplify the later analysis process.

@@

The Data Carpentry
lesson on using spreadsheets (see <http://bit.ly/datacarp_scm>)
has a nice table with good and bad example variable names:

**good name**       | **good alternative** | **avoid**
------------------- | -------------------- | ---------
`Max_temp`          | `MaxTemp1`           | `Maximum Temp (°C)`
`Precipitation`     | `Precipitation_mm`   | `precmm`
`Mean_year_growth`  | `MeanYearGrowth`     | `Mean growth/year`
`sex`               | `sex`                | `M/F`
`weight`            | `weight`             | `w.`
`cell_type`         | `CellType`           | `Cell type`
`first_observation` | `Observation_01`     | `1st Obs.`

@@

But note that Microsoft Excel does bizarre things with dates
(see <https://storify.com/kara_woo/excel-date-system-fiasco>). It stores them
internally as an number, counting the days since 1900-01-01. Wait, that is only
for Windows; on Macs, it counts the days since 1904-01-01. So, you may
need to manually check that the dates haven't been mangled when your data
come out of Excel.

@@

The best layout for your data within in a spreadsheet is as a single big
rectangle with rows corresponding to subjects and columns corresponding to
variables. The first row should contain variable names.
(*Please don't use more than one row for the variable names.*)
An example of a rectangular layout is shown in Figure 4.

@@

Some data sets won't fit nicely into a single rectangle, but they will usually
fit into a set of rectangles, in which case you can make a set of Excel files,
each with a rectangle of data. It is best to keep each rectangle in its own
file; tables scattered around a worksheet are difficult to work with, and they
make it hard to export the data to CSV files (which we will discuss shortly).
You might also consider having a single Excel file with multiple
worksheets. We prefer to have multiple files with one sheet each so we
can more easily save the data as CSV files, but if you do use multiple
worksheets in a file be sure to use a consistent structure.

@@

Alternatively, make it a "tidy" data set
with each row being a subject on a
particular day, as shown in Figure 8.

@@

An example data dictionary is displayed in Figure 9.
Note that this is a rectangular data set, like any other. The first
column contains the variable names. The second column is a more readable
version, as might be used in data visualizations. The third column
groups the variables into different categories, which might
also be used in data visualizations. The last column is a description.

@@

For example, in Figure 10A, a suspicious entry is highlighted. It would be
better to include an additional column that indicates the outliers (as in Figure
10B). The highlighting is nice visually, but it is hard to extract that
information for use in the later analysis. Analysis programs can much more
readily handle data that is stored in a column than that which is encoded in
cell highlighting, font, etc. (and in fact this markup will be lost completely
in many programs).

>>>>

Murrell [-@murrell2013] contrasts data that are formatted for humans to
view by eye with data that are formatted for a computer. He provides an
extended example of computer code to extract data from a set of files
with complex arrangements. It is important that data analysts be able
to work with such complex data files. But if the initial arrangement
of the data files is planned with the computer in mind, this would
simplify the later analysis process.

@@

The Data Carpentry
lesson on using spreadsheets (see <http://www.datacarpentry.org/spreadsheet-ecology-lesson/02-common-mistakes>)
has a nice table with good and bad example variable names:

**good name**       | **good alternative** | **avoid**
------------------- | -------------------- | ---------
`Max_temp_C`        | `MaxTemp`            | `Maximum Temp ($^{\circ}$C)`
`Precipitation_mm`  | `Precipitation`      | `precmm`
`Mean_year_growth`  | `MeanYearGrowth`     | `Mean growth/year`
`sex`               | `sex`                | `M/F`
`weight`            | `weight`             | `w.`
`cell_type`         | `CellType`           | `Cell type`
`Observation_01`    | `first_observation`  | `1st Obs.`

@@

But note that Microsoft Excel does bizarre things with dates
(see <https://storify.com/kara_woo/excel-date-system-fiasco>). It stores them
internally as a number, counting the days since 1900-01-01. Wait, that is only
for Windows; on Macs, it counts the days since 1904-01-01. So, you may
need to manually check that the dates haven't been mangled when your data
come out of Excel.

@@

The best layout for your data within a spreadsheet is as a single big
rectangle with rows corresponding to subjects and columns corresponding to
variables. The first row should contain variable names.
(*Please don't use more than one row for the variable names.*)
An example of a rectangular layout is shown in Figure 4.

@@

Some datasets won't fit nicely into a single rectangle, but they will usually
fit into a set of rectangles, in which case you can make a set of Excel files,
each with a rectangle of data. It is best to keep each rectangle in its own
file; tables scattered around a worksheet are difficult to work with, and they
make it hard to export the data to CSV files (which we will discuss shortly).
You might also consider having a single Excel file with multiple
worksheets. We prefer to have multiple files with one sheet each so we
can more easily save the data as CSV files, but if you do use multiple
worksheets in a file be sure to use a consistent structure.

@@

Alternatively, make it a "tidy" dataset
with each row being a subject on a
particular day, as shown in Figure 8.

@@

An example data dictionary is displayed in Figure 9.
Note that this is a rectangular dataset, like any other. The first
column contains the variable names. The second column is a more readable
version, as might be used in data visualizations. The third column
groups the variables into different categories, which might
also be used in data visualizations. The last column is a description.

@@

For example, in Figure 10A, a suspicious entry is highlighted. It would be
better to include an additional column that indicates the outliers (as in Figure
10B). The highlighting is nice visually, but it is hard to extract that
information for use in the later analysis. Analysis programs can much more
readily handle data that are stored in a column than that which is encoded in
cell highlighting, font, etc. (and in fact this markup will be lost completely
in many programs).
